---
execute: 
  echo: false
---


```{r}
#| message: false
library(tidyverse)

x <- read_csv("data/nyc_tracts.csv")
```


```{r}
#| message: false
# Load required libraries for data manipulation and visualization
library(ggplot2)  # For creating beautiful and customizable plots
library(dplyr)    # For data manipulation and transformation
library(scales)   # For formatting axes labels (currency, percentages, etc.)
library(stringr)  # For string manipulation and pattern matching

# DATA PREPARATION AND CLEANING
# The raw data contains census tract names with borough information embedded in the tract description
# We need to extract this geographic information and standardize race categories for analysis

# Step 1: Clean the dataset and extract meaningful geographic and demographic information
x_clean <- x %>%
  # Remove census tracts with missing income data to ensure accurate analysis
  filter(!is.na(med_income)) %>%
  # Extract borough names from the tract description string
  # This allows us to analyze income patterns by NYC borough
  mutate(
    borough = case_when(
      str_detect(tract, "Queens County") ~ "Queens",
      str_detect(tract, "Kings County") ~ "Brooklyn", 
      str_detect(tract, "New York County") ~ "Manhattan",
      str_detect(tract, "Bronx County") ~ "Bronx",
      str_detect(tract, "Richmond County") ~ "Staten Island",
      TRUE ~ "Other"
    )
  ) %>%
  # Standardize race/ethnicity categories for consistent analysis
  # This handles variations in how racial categories might be coded
  mutate(
    race = case_when(
      str_detect(race, "White") ~ "White",
      str_detect(race, "Black") ~ "Black",
      str_detect(race, "Hisp") ~ "Hispanic",
      str_detect(race, "Asian") ~ "Asian",
      TRUE ~ race
    )
  )

# VISUALIZATION 1: INCOME DISTRIBUTION BY RACE ACROSS NYC BOROUGHS
# This plot reveals how income varies both by race and geography within New York City
# Box plots show the median, quartiles, and outliers, while jittered points show individual tract data

p1 <- ggplot(x_clean, aes(x = race, y = med_income, fill = race)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  geom_jitter(alpha = 0.3, width = 0.2, size = 0.5) +
  facet_wrap(~borough, scales = "free_x") +
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  scale_fill_viridis_d(name = "Race/Ethnicity") +
  labs(
    title = "Median Income Distribution by Race Across NYC Boroughs",
    subtitle = "Distribution of median household income by census tract",
    x = "Race/Ethnicity",
    y = "Median Household Income",
    caption = "Data: Census tract-level median income by race\nNote: Each point represents a census tract"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

# SUMMARY STATISTICS TABLE (for internal use - not displayed)
# Calculate key statistics to quantify income disparities across racial and geographic lines
summary_stats <- x_clean %>%
  group_by(race, borough) %>%
  summarise(
    count = n(),
    median_income = median(med_income, na.rm = TRUE),
    mean_income = mean(med_income, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(borough, desc(median_income))

# Display the first visualization


# VISUALIZATION 2: OVERALL INCOME DISTRIBUTION ACROSS ALL NYC
# This violin plot shows the shape of income distribution for each racial group
# Wider sections indicate more census tracts at that income level

p2 <- ggplot(x_clean, aes(x = reorder(race, med_income, median), y = med_income, fill = race)) +
  geom_violin(alpha = 0.7) +    # Shows distribution shape/density
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.alpha = 0.5) +  # Shows quartiles and outliers
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  scale_fill_viridis_d(name = "Race/Ethnicity") +
  labs(
    title = "Overall Income Distribution by Race in NYC",
    subtitle = "Violin plots showing income distribution density",
    x = "Race/Ethnicity (ordered by median income)",
    y = "Median Household Income"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  ) +
  coord_flip()  # Horizontal orientation for better readability



# VISUALIZATION 3: BOROUGH-LEVEL INCOME COMPARISON
# This bar chart directly compares median incomes across boroughs for each racial group
# It helps identify which boroughs offer better economic opportunities for different communities

borough_summary <- x_clean %>%
  group_by(borough, race) %>%
  summarise(
    median_income = median(med_income, na.rm = TRUE),
    count = n(),
    .groups = 'drop'
  ) %>%
  filter(count >= 5)  # Only include combinations with sufficient data (at least 5 census tracts)

p3 <- ggplot(borough_summary, aes(x = borough, y = median_income, fill = race)) +
  geom_col(position = "dodge", alpha = 0.8) +  # Side-by-side bars for easy comparison
  scale_y_continuous(labels = dollar_format(scale = 1e-3, suffix = "K")) +
  scale_fill_viridis_d(name = "Race/Ethnicity") +
  labs(
    title = "Median Income by Race and Borough",
    subtitle = "Comparing median household income across NYC boroughs",
    x = "Borough",
    y = "Median Household Income"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom"
  )


```


## Income Distribution by Race Across NYC Boroughs

This visualization reveals how income varies both by race and geography within New York City. The box plots show the median, quartiles, and outliers for each racial group, while the individual points represent census tracts. Each panel shows a different borough, allowing us to compare income patterns across NYC's diverse neighborhoods.

```{r}
#| echo: false
# Your first plot code here
p1
```
## Overall Income Distribution by Race
This violin plot provides a city-wide perspective on income inequality across racial groups. The violin shapes show the density of census tracts at different income levels - wider sections indicate more common income ranges. Racial groups are ordered from lowest to highest median income, making the disparities clearly visible.


```{r}
p2
```

## Borough-Level Income Comparison
This bar chart enables direct comparison of median incomes across boroughs for each racial group. It reveals whether income gaps exist consistently across all boroughs or vary by geographic location. Only racial groups with sufficient data (5+ census tracts) in each borough are shown to ensure statistical reliability.


```{r}
p3
```
## Key Findings
These three visualizations together provide a comprehensive view of income inequality in NYC, revealing patterns of economic segregation and opportunity across the city's diverse communities.

